---
date: 'Nov. - Dec. 2020'
output: 
  github_document:
    html_preview: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
set.seed(2018)
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)

theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
tri_colors <-  c("darkorange","darkorchid","cyan4") # https://allisonhorst.github.io/palmerpenguins/articles/intro.html
mutate_perf <- function(dat) {
  dat %>% 
    mutate(mem_mb = max_rss,
           mem_gb = mem_mb / 1024) %>% 
    rename(sec = s,
           num_otus = sobs)
}
```

## _de novo_ clustering

```{r load_opticlust}
opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
  mutate_perf()
sum_opticlust <- opticlust %>% 
  group_by(dataset) %>% 
  summarize(mean_mcc = mean(mcc),
            sd_mcc = sd(mcc),
            mean_sec = mean(sec),
            sd_sec = sd(sec),
            mean_mem_gb = mean(mem_gb))#,
            #mean_shannon = mean(shannon))
```

## fit to reference databases

```{r load_fit-dbs}
ref_sizes <- read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
  mutate_perf()
head(optifit_dbs)
```

Plotted the fraction mapped for open-reference clustering too as a sanity check.
Final plots to be included in the paper won't include open-ref.
```{r fraction-mapped-open_fit-db}

optifit_dbs %>% #filter(method == 'closed') %>% 
  ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
  geom_boxplot() + 
  facet_wrap('method') +
  ylim(0, 1) +
  scale_color_manual(values=tri_colors) +
  labs(title="Fit to reference database") +
  theme(plot.caption = element_markdown())
```

## fit split datasets

Using buggy data from commit `824c1306eb464349f9d331f85550d539f8c66c63`.

```{bash, eval = FALSE}
wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/824c1306eb464349f9d331f85550d539f8c66c63/subworkflows/3_fit_sample_split/results/optifit_split_results.tsv?token=AEHR6TKMLXIG57IGER2VXSS74C5WC -O exploratory/2020-11_Nov-Dec/debug-split_100-seeds.tsv
```


```{r load_fit-split}
optifit_split <- read_tsv(here('exploratory',
                               '2020-11_Nov-Dec',
                               'debug-split_100-seeds.tsv')) %>% 
  mutate_perf()
```

```{r debug-mcc_fit-split}
optifit_split %>% 
  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
  facet_grid(dataset ~ method) +
  ylim(0, 1) +
  labs(title='OTU Quality: OptiFit with split datasets',
       x='reference fraction',
       caption='Black line: _de novo_ clustering on the whole dataset') +
  theme(plot.caption = element_markdown())
```

### fraction mapped

```{r debug-fraction-mapped_fit-split}
optifit_split %>% 
  filter(method == 'closed', tool == 'mothur', is.na(ref)) %>% 
  #group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
  geom_point(alpha = 0.5, size=1) +
  facet_wrap("dataset", nrow=1) +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
```

It looks like one result per parameter set is left over from before I re-ran 
everything after reverting to the older silva version.
[Here's what the plot looked like with the newest (buggy) silva](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/exploratory/2020-05/sub3_fit_all-seqs.md#fraction-of-sequences-that-map-to-the-reference).
Need to go back and figure out why those files didn't get removed/overwritten
by Snakemake when I re-ran everything.

Let's try a sanity check with open-reference -- should all be 1.

```{r debug-fraction-mapped-open_fit-split}
optifit_split %>% 
  filter(tool == 'mothur', is.na(ref)) %>% 
  #filter(dataset == 'soil') %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
  geom_point(alpha = 0.5, size=1) +
  facet_grid(method ~ dataset) +
  ylim(0, 1) +
  labs(title="Fit dataset splits",
       x='reference fraction',
       y='fraction mapped')
```

Did mothur's list file format change in the latest dev version? The fraction
mapped values are the same as the reference fraction when the ref weight is
simple.

### is there a pattern?

```{r debug-pattern}
debug_dat <- optifit_split %>% 
  filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, method, fraction_mapped, is_correct, 
         sample_frac, ref_frac, ref_weight)

debug_dat %>% filter(is_correct)

```

Can't discern any pattern...

## smaller dataset

Using buggy data from commit `1de7b839eae9b222b66f15dd24b0bb54a8a35459`.

```{bash, eval = FALSE}
wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/1de7b839eae9b222b66f15dd24b0bb54a8a35459/subworkflows/3_fit_sample_split/results/optifit_split_results.tsv?token=AEHR6TJQFZT4XT6XIYWOKNS74C7ME -O exploratory/2020-11_Nov-Dec/debug-split_2-seeds.tsv
```

```{r debug-2_fit-split}
split_2 <- read_tsv(here('exploratory',
                          '2020-11_Nov-Dec',
                          'debug-split_2-seeds.tsv')) %>%
  mutate_perf() %>% 
  filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, method, fraction_mapped, is_correct, 
         sample_frac, ref_frac, ref_weight)

split_2 %>% 
  filter(dataset == 'soil', ref_weight == 'simple') %>% 
  select(sample_frac, ref_frac, seed, is_correct) %>% 
  pivot_wider(names_from = seed, 
              values_from = is_correct,
              names_glue = 'seed_{seed}_is_correct')

```

The pattern of correctness among ref fracs is flipped between seed 1 and 2

Whenever the fit seed matches the best cluster seed,
the fraction mapped results are correct.
The bug disappears if you only use 1 seed, because then
all fit runs will use the same seed as the data split.