---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---

# Similarity & Abundance of Sequences

Each sequence has an absolute abundance and similarity value, where similarity
is the count of other sequences within the dataset that it is similar to per the
0.03 distance threshold. I was wondering if the distributions of those values
would have some pattern that corresponded to what we see with the MCC scores for
weighting by those methods. There isn't anything obviously weird or wrong. For
all datasets most sequences are very low abundance (expected). Both marine and
soil have a much lower median similarity value than human and mouse, and the
marine and soil datasets also have a huge spread in MCC between open & closed
with similarity weighting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/",
                      dpi = 300
                      )
```
```{r deps, warning=FALSE, message=FALSE}
set.seed(20211021)
library(glue)
library(here)
library(tidyverse)
theme_set(theme_bw())
```

```{r dat}
dataset_names <- c("human","marine","mouse","soil")
seq_stats <- dataset_names %>% map_dfr(function(x) {
  read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
    mutate(dataset = x)
}) %>% pivot_longer(c(similarity, abundance), names_to = 'statistic')
```

### histograms

```{r hist_sim_abun}
seq_stats %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  facet_grid(dataset ~ statistic, scales='free')
```

```{r hist_sim_abun_log10}
seq_stats %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  scale_x_log10() +
  facet_wrap("dataset")
```
```{r hist_sim_abun_log10bins}
seq_stats %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram(bins=10) +
  scale_x_log10() +
  facet_grid(dataset ~ statistic, scales='free')
```

### boxplots

```{r box_sim_abun}
seq_stats %>% 
  ggplot(aes(x=value, y=dataset, fill=statistic)) +
  geom_boxplot() +
  facet_wrap('statistic', scales = 'free')
``` 
```{r box_sim_abun_log10}
seq_stats %>% 
  ggplot(aes(x=value, y=dataset, fill=statistic)) +
  geom_boxplot() +
  scale_x_log10() +
  facet_wrap('statistic', scales = 'free')
``` 


### abundance vs similarity

```{r point_abun_sim}
seq_stats %>% 
  pivot_wider(names_from = statistic) %>% 
  ggplot(aes(x = abundance, y = similarity, color = dataset)) +
  geom_point(alpha=0.7) +
  scale_x_log10() +
  facet_wrap('dataset')
```


## relative abundance & simlarity

```{r calc_relative_stats}
totals <- seq_stats %>%
  pivot_wider(names_from = statistic) %>% 
  group_by(dataset) %>% 
  summarize(tot_simi = sum(similarity),
            tot_abun = sum(abundance))

rel_stats <- dataset_names %>% map_dfr(function(x) {
  read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
    mutate(dataset = x,
           rel_sim = similarity / (totals %>% 
                                     filter(dataset == x) %>% 
                                     pull(tot_simi)),
           rel_abun = abundance / (totals %>% 
                                     filter(dataset == x) %>% 
                                     pull(tot_abun))
           )
})
```

### rel abun vs rel sim

```{r point_rel_abun_sim}
rel_stats %>% 
  ggplot(aes(x = rel_abun, y = rel_sim, color = dataset)) +
  geom_point(alpha=0.7) +
  facet_wrap('dataset')
```
```{r point_rel_abun_sim_log10}
rel_stats %>% 
  ggplot(aes(x = rel_abun, y = rel_sim, color = dataset)) +
  geom_point(alpha=0.7) +
  scale_x_log10() +
  facet_wrap('dataset')
```

### relative boxplots

```{r box_rel_sim_abun}
rel_stats %>% pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
  ggplot(aes(x=value, y=dataset, fill=statistic)) +
  geom_boxplot() +
  facet_wrap('statistic', scales = 'free')
``` 


```{r box_rel_sim_abun_log10}
rel_stats %>% pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
  ggplot(aes(x=value, y=dataset, fill=statistic)) +
  geom_boxplot() +
  scale_x_log10() +
  facet_wrap('statistic', scales = 'free')
```

### relative histograms

```{r hist_rel_sim_abun}
rel_stats %>%  pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  facet_wrap("dataset")
```

```{r hist_rel_sim_abun_log10}
rel_stats %>%  pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  scale_x_log10() +
  facet_wrap("dataset")
```

## summary stats

```{r summary_abundance}
rel_stats %>% 
  group_by(dataset) %>% 
  summarize_at(vars(abundance), 
               list(min=min, median=median, mean=mean, max=max)) %>% 
  knitr::kable()
```


```{r summary_similarity}
rel_stats %>% 
  group_by(dataset) %>% 
  summarize_at(vars(similarity), 
               list(min=min, median=median, mean=mean, max=max)) %>% 
  knitr::kable()
```

```{r quantiles}
quantile_custom <- function(x, by = 0.05) {
  x %>% 
    quantile(., seq.int(from = 0.1, to = 1, by = by))
}
get_quantile <- function(dat, dataset_name, var = abundance) {
  quantiles <- dat %>% 
    filter(dataset == dataset_name) %>% 
    pull({{ var }}) %>%  
    quantile_custom()
  data.frame(quant = names(quantiles), 
             value = quantiles, 
             row.names = NULL) %>% 
    mutate(dataset = dataset_name)
}
get_dataset_quantiles <- function(dat, column = abundance) {
  dataset_names %>% 
    map_dfr(function(x) {get_quantile(dat = dat, 
                                    dataset_name = x,
                                    var = {{column}})
                         }) %>% 
    pivot_wider(names_from = dataset, values_from = value) 
}
```
 
### abundance quantiles
 
```{r abun_quant}
rel_stats %>% 
  get_dataset_quantiles(column = abundance) %>% 
  knitr::kable()
```

### similarity quantiles
 
```{r simi_quant}
rel_stats %>% 
  get_dataset_quantiles(column = similarity) %>% 
  knitr::kable()
```

