---
title: "OptifitLabMeetingPlots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

## Generation of figures for end of semester rotation presentation

This is the code used to generate the figures presented at lab meeting on 12/12/18. The data was generated from the last version of the bash scripts, not from snakemake. The sensspec file here was generated automatically, except for the final 60 lines containing purely OptiClust MCC data which I added manually for a 100% of data OptiClust comparison.

For all plots here, a single point is the MCC score from a single run of OptiFit or OptiClust.

First we import the data and create factors (so things don't just plot alphabetically) and full text labels for later use in our plots.

```{r cars}
mice_labmeeting <- read_tsv("mice.sensspec.labmeeting")
mice_labmeeting$method <- factor(mice_labmeeting$method, levels=c("none", "samp.abund", "samp.dist",
                                                                    "ref.dist", "ref.abund", "OptiClust"))
mice_labmeeting$type <- factor(mice_labmeeting$type, levels = c("REF", "SAMP",
                                                                  "SAMP_C_REF", "SAMP_O_REF", "SAMP_C_NOREF" ,"SAMP_O_NOREF"))
mice_labmeeting$label <- sapply(mice_labmeeting$type, function(x)  c("REF", "SAMP", "Closed, with ref",
                                                       "Open, with ref", "Closed, no ref", "Open, no ref")[x])
```

The first comparison plot we want to make is for the reference alone using OptiClust. This just lets us see how the MCC score varies using different sized datasets. As noted in the legend, the weighting method for sampling here is "none" meaning sequences were completely randomly chosen to be a part of the reference. We can see that the MCC seems to converge as the number of sequences increases.

```{r mice_ref}
mice_ref <- filter(mice_labmeeting, method %in% c("none") & type %in% c("REF"))

mice_ref_plot <- ggplot(data=mice_ref, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_ref_plot 
```


Next is a comparison of weighting methods when choosing sequences to go into the reference. First we look at random sampling vs sampling by abundance. We can see that in almost every case, MCC scores obtained when sampling by abundance are superior to sampling completely randomly.
```{r mice_abund}
mice_by_abund <- filter(mice_labmeeting, method %in% c("ref.abund", "none") & !(type %in% c("REF", "SAMP")))
mice_abund_plot <- ggplot(data=mice_by_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_abund_plot
```

We also want to look at weighting our reference selection by the number of pairwise connections (distances <3%) per sequence. Here we can see that when dropping the reference from the final output, weighting by connections gives lower MCC scores, but when including the reference it appears to give higher MCC scores when compared to completely random sampling.

```{r mice_dist}
mice_by_dist <- filter(mice_labmeeting, method %in% c("ref.dist", "none") & !(type %in% c("REF", "SAMP")))
mice_dist_plot <- ggplot(data=mice_by_dist, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_dist_plot
```

Now for a direct comparison of sampling by abundance and sampling by connections. Except in the case of small references when closed fitting with reference, sampling by abundance produces consistently higher MCC scores.

```{r mice_dist_abund}
mice_by_dist_abund <- filter(mice_labmeeting, method %in% c("ref.dist", "ref.abund") & !(type %in% c("REF", "SAMP")))
mice_dist_abund_plot <- ggplot(data=mice_by_dist_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_dist_abund_plot
```

To look at random variation, we look back at our first plot when sampling by abundance, but look at the individual data points. We can see that while the MCC scores generally cluster to a single value, sometimes they randomly drop out to a lower MCC. Overall this effect seems to be rare though, especially when sampling by abundance. Statistical quantification of this "drop-out" rate may be necessary.

```{r abund_variation}
mice_abund_variation_plot <- ggplot(data=mice_by_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_abund_variation_plot
```

Finally, we wanted to compare OptiFit's results to OptiClust to ensure that the algorithm tends to give similar MCC scores in similar scenarios. This helps verify that the algorithm is doing what we think it should be doing, and gives confidence that we are getting MCC scores as close to optimal as possible. We can see that OptiFit does indeed give MCC scores almost equivalent to OptiClust, with most scores be ~.1% less than OptiClust. Even the random "drop-outs" are only around 2% worse than OptiClust, indicating overall good MCC scores when comparing the two methods.

```{r opticlust}
mice_opticlust <- filter(mice_labmeeting, method %in% c("ref.abund", "OptiClust") & !(type %in% c("REF", "SAMP",
                                                                                                   "SAMP_C_REF", "SAMP_C_NOREF",
                                                                                                   "SAMP_O_NOREF")))
opticlust_plot <- ggplot(data=mice_opticlust, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

opticlust_plot
```

